Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2014-11-11 MS_revised.tex
moved BAMM text to discussion
Fix Satyrini, it should be Satyrinae in many cases

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrini (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table 1).

@@

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the splits in the MCC tree are recovered in very few trees, for
example the split for the clade Satyrini (Euptychiina + Pronophilina +
Satyrina + Maniolina) is recovered with a probability of 0.18 (Fig. 3).

@@

In this study, the effect of phylogenetic uncertainty on the inferred
diversification splits by MEDUSA is amplified because some Nymphalidae
taxa appear to be strongly affected by long-branch attraction artifacts
\cite{pena2011}. Thus, the Bayesian runs are expected to recover alternative
topologies on the posterior distribution of trees, resulting in low
support and posterior probability values for the nodes. For example,
posterior probability values for clades in Satyrini are very low
\cite{wahlberg2009}. As a result, MEDUSA inferred a net diversification rate
increase for part of Satyrini in the maximum clade credibility tree, but
this was recovered in the MultiMEDUSA analysis in only 13\% of the
random sample of trees.

@@

@@

\subsubsection*{Satyrini}

The diverse tribe Satyrini radiated simultaneously with the radiation of
their main hostplant, grasses, during the climatic cooling in the
Oligocene \cite{pena2008}. Thus, it is somewhat surprising that part of
Satyrini (the subtribes Euptychiina, Satyrina and Pronophilina) were
found to have accelerated diversification in only 13\% of the trees from
the posterior distribution. Although this can be attributed to low
phylogenetic signal \cite{wahlberg2009}, the clade Satyrini is very robust
\cite{wahlberg2009}
and MEDUSA failed to identify any significant accelerated net
diversification rate for Satyrini. It appears that the radiation of
Satyrini as a whole was not remarkably fast and therefore not picked up
by MEDUSA. This should be expected if the diversification of Satyrinae
occurred in a stepwise manner, with pulses or bursts of diversification
for certain lineages but unlikely for the tribe Satyrini as a whole.

\section*{Conclusions}
We found that even though MEDUSA estimated several diversification
shifts in the maximum clade credibility tree of Nymphalidae, only a few
of these splits were found in more than 90\% of the trees from the
posterior distribution. In the literature, it is common practice that
conclusions are based on the splits estimated on the maximum clade
credibility tree. However, by using a MultiMEDUSA approach, we found
that for this Nymphalidae dataset some of these splits might be greatly
affected by phylogenetic uncertainty. Moreover, some of these splits can
be recovered either as increases or decreases in net diversification
rate depending on the tree from the posterior distribution that was used
for analysis. This means that contradictory conclusions would be made if
only the maximum clade credibility tree was used for analysis. We
recommend that all datasets should be analyzed using both approaches,
MEDUSA and MultiMEDUSA, in order to test whether the results are robust
when phylogenetic uncertainty is taken into account.
The results from BAMM recovered only two significant shifts in our MCC tree
as this method is more conservative than the MEDUSA approach. One of these 
shifts is the Satyrini clade, which was recovered by MEDUSA in a few trees 
from the posterior distribution. Thus, it might be important to
take into account phylogenetic uncertainty in BAMM as well.

@@

\section*{Figure legends}
\begin{description}
 \item {\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrini, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicate the posterior probability values for
such nodes.

@@

% Table 1
\section*{Tables}
\begin{table}[!h]
\caption{\bf{Significant net diversification rate shifts found in the MEDUSA analysis of the Nymphalid maximum clade credibility tree.}}
\begin{tabular}{lccrrcl}
Shift N$^\circ$ & Split.Node & Model & r          & LnLik.part & AICc     & Taxa                                                       \\
1               & 399        & yule  & 0.092459   & -1055.957  &          & Nymphalidae (root)                                         \\
2               & 691        & bd    & 0.054129   & -406.3703  &          & Limenitidinae + Heliconiinae                               \\
3               & 299        & yule  & 0.311199   & -6.3058    &          & \emph{Ypthima}                                             \\
4               & 224        & yule  & 0.290989   & -6.2601    &          & \emph{Charaxes}                                            \\
5               & 750        & yule  & 0.186913   & -147.4146  &          & Oleriina + Ithomiina + Napeogenina + Dircennina + Godyrina \\
6               & 405        & yule  & 0.116252   & -555.0276  &          & Satyrini                                                   \\
7               & 495        & yule  & 0.064656   & -124.9143  &          & Coenonymphina                                              \\
8               & 609        & yule  & 0.240562   & -35.0908   &          & Phyciodina in part                                         \\
9               & 787        & bd    & 0.042332   & -43.7819   &          & Danaini in part                                            \\
10              & 231        & yule  & 0.209416   & -4.7955    &          & \emph{Coenonympha}                                         \\
11              & 478        & yule  & 0.311684   & -9.2218    &          & \emph{Caeruleuptychia} + \emph{Magneuptychia}              \\
12              & 659        & yule  & 0.219253   & -11.2686   &          & \emph{Callicore} + \emph{Diaethria}                        \\
13              & 444        & yule  & 0.220615   & -43.7812   &          & Satyrina                                                   \\
14              & 524        & yule  & 0.190754   & -26.0651   &          & Mycalesina                                                 \\
15              & 355        & yule  & 0.234041   & -6.2013    &          & \emph{Pedaliodes}                                          \\
16              & 714        & yule  & 0          & 0          &          & \emph{Dryas} + \emph{Dryadula}                             \\
17              & 377        & yule  & 0.311671   & -4.1986    &          & \emph{Taenaris}                                            \\
18              & 688        & yule  & 0.024724   & -17.5256   &          & Pseudergolinae                                             \\
19              & 583        & yule  & 0          & 0          & 5090.492 & \emph{Anaeomorpha} + \emph{Hypna}                                       
\end{tabular}
\begin{flushleft}Split.Node=node number, Model=preferred diversification model by MEDUSA, r=net diversification rate, LnLik.part=log likelihood value.
\end{flushleft}
\end{table}

>>>>

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrinae (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table 1).

@@

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the splits in the MCC tree are recovered in very few trees, for
example the split for the clade Satyrinae is recovered with a probability
of 0.18 (Fig. 3).

@@

In this study, the effect of phylogenetic uncertainty on the inferred
diversification splits by MEDUSA is amplified because some Nymphalidae
taxa appear to be strongly affected by long-branch attraction artifacts
\cite{pena2011}. Thus, the Bayesian runs are expected to recover alternative
topologies on the posterior distribution of trees, resulting in low
support and posterior probability values for the nodes. For example,
posterior probability values for clades in Satyrinae are very low
\cite{wahlberg2009}. As a result, MEDUSA inferred a net diversification rate
increase for the Satyrinae (which includes Satyrini and its sister clade)
in the maximum clade credibility tree, but
this was recovered in the MultiMEDUSA analysis in only 13\% of the
random sample of trees.

@@

The results from BAMM recovered only two significant shifts in our MCC tree
as this method seems to be much more conservative than the MEDUSA approach.
One of these 
shifts is the Satyrini clade. MEDUSA estimated its parent node as
significant shift in a few trees from the posterior distribution. 
Thus, it might be important to
take into account phylogenetic uncertainty in BAMM as well.

@@

\subsubsection*{Satyrinae}

Lineages in the diverse family Satyrinae radiated simultaneously with the
radiation of
their main hostplant, grasses, during the climatic cooling in the
Oligocene \cite{pena2008}. Thus, it is somewhat surprising that part of
Satyrinae were
found to have accelerated diversification in only 13\% of the trees from
the posterior distribution. Although this can be attributed to low
phylogenetic signal \cite{wahlberg2009}, the clade Satyrini is very robust
\cite{wahlberg2009}
and MEDUSA failed to identify any significant accelerated net
diversification rate for Satyrini. It appears that the radiation of
Satyrini as a whole was not remarkably fast and therefore not picked up
by MEDUSA, although it estimated a diversification shift for Satyrini + its
sister clade. This should be expected if the diversification of Satyrini
occurred in a stepwise manner, with pulses or bursts of diversification
for certain lineages but unlikely for the tribe Satyrini as a whole.

\section*{Conclusions}
We found that even though MEDUSA estimated several diversification
shifts in the maximum clade credibility tree of Nymphalidae, only a few
of these splits were found in more than 90\% of the trees from the
posterior distribution. In the literature, it is common practice that
conclusions are based on the splits estimated on the maximum clade
credibility tree. However, by using a MultiMEDUSA approach, we found
that for this Nymphalidae dataset some of these splits might be greatly
affected by phylogenetic uncertainty. Moreover, some of these splits can
be recovered either as increases or decreases in net diversification
rate depending on the tree from the posterior distribution that was used
for analysis. This means that contradictory conclusions would be made if
only the maximum clade credibility tree was used for analysis. We
recommend that all datasets should be analyzed using both approaches,
MEDUSA and MultiMEDUSA, in order to test whether the results are robust
when phylogenetic uncertainty is taken into account.

@@

\section*{Figure legends}
\begin{description}
 \item {\bf Figure 1. Results of the MEDUSA analysis run on the maximum
clade credibility tree from Wahlberg et al. (2009)} Rate shifts were estimated for the
following nodes (besides the background rate): 2) Limenitidinae +
Heliconiinae, 3) \emph{Ypthima}, 4) \emph{Charaxes}, 5) Ithomiini in
part, 6) Satyrinae, 7) Coenonymphina 8) Phyciodina in part, 9) Danaini in
part, 10) \emph{Coenonympha}, 11) \emph{Caeruleuptychia} +
\emph{Magneuptychia}, 12) \emph{Callicore} + \emph{Diaethria}, 13)
Satyrina, 14) Mycalesina, 15) \emph{Pedaliodes}, 16) \emph{Dryas} +
\emph{Dryadula}, 17) \emph{Taenaris}, 18) Pseudergolinae. Circles on
nodes indicate the diversification shift number as found by MEDUSA.
Numbers next to circles indicate the posterior probability values for
such nodes.

@@

% Table 1
\section*{Tables}
\begin{table}[!h]
\caption{\bf{Significant net diversification rate shifts found in the MEDUSA analysis of the Nymphalid maximum clade credibility tree.}}
\begin{tabular}{lccrrcl}
Shift N$^\circ$ & Split.Node & Model & r          & LnLik.part & AICc     & Taxa                                                       \\
1               & 399        & yule  & 0.092459   & -1055.957  &          & Nymphalidae (root)                                         \\
2               & 691        & bd    & 0.054129   & -406.3703  &          & Limenitidinae + Heliconiinae                               \\
3               & 299        & yule  & 0.311199   & -6.3058    &          & \emph{Ypthima}                                             \\
4               & 224        & yule  & 0.290989   & -6.2601    &          & \emph{Charaxes}                                            \\
5               & 750        & yule  & 0.186913   & -147.4146  &          & Oleriina + Ithomiina + Napeogenina + Dircennina + Godyrina \\
6               & 405        & yule  & 0.116252   & -555.0276  &          & Satyrinae                                                   \\
7               & 495        & yule  & 0.064656   & -124.9143  &          & Coenonymphina                                              \\
8               & 609        & yule  & 0.240562   & -35.0908   &          & Phyciodina in part                                         \\
9               & 787        & bd    & 0.042332   & -43.7819   &          & Danaini in part                                            \\
10              & 231        & yule  & 0.209416   & -4.7955    &          & \emph{Coenonympha}                                         \\
11              & 478        & yule  & 0.311684   & -9.2218    &          & \emph{Caeruleuptychia} + \emph{Magneuptychia}              \\
12              & 659        & yule  & 0.219253   & -11.2686   &          & \emph{Callicore} + \emph{Diaethria}                        \\
13              & 444        & yule  & 0.220615   & -43.7812   &          & Satyrina                                                   \\
14              & 524        & yule  & 0.190754   & -26.0651   &          & Mycalesina                                                 \\
15              & 355        & yule  & 0.234041   & -6.2013    &          & \emph{Pedaliodes}                                          \\
16              & 714        & yule  & 0          & 0          &          & \emph{Dryas} + \emph{Dryadula}                             \\
17              & 377        & yule  & 0.311671   & -4.1986    &          & \emph{Taenaris}                                            \\
18              & 688        & yule  & 0.024724   & -17.5256   &          & Pseudergolinae                                             \\
19              & 583        & yule  & 0          & 0          & 5090.492 & \emph{Anaeomorpha} + \emph{Hypna}                                       
\end{tabular}
\begin{flushleft}Split.Node=node number, Model=preferred diversification model by MEDUSA, r=net diversification rate, LnLik.part=log likelihood value.
\end{flushleft}
\end{table}
